Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-03-17 paper.tex
update

The abstract syntax can be straightforwardly described in Scala as follows:
\note{TR: grammar below isn't ANF. Either make it so, or clarify that
we assume ANF only as convention and don't need/want to enforce it in
the term structure}

@@

The CESK machine is an
abstract machine for describing semantics of and evaluating $\lambda$-calculus \cite{felleisen1987calculus}.
The CESK machine models program execution as state transitions in a small-step fashion.
As its name suggests, a machine state has four components:
1) \textit{Control} is the expression currently being evaluated.
2) \textit{Environment} is a map that contains the address of a variable in the lexical scope.
3) \textit{Store} models the heap of a program as a map from addresses to values.
The address space consists of numbers (0-indexed).
In our toy language, the only category of value is a closure, i.e., a function paired with an environment.
4) \textit{Continuation} represents the program stack. \todo{explain more on stack}

The Scala representations for the components of CESK machine are as follows:

@@

It is worth noting that the continuation class @Kont@ is defined as a list of
frames, where a frame can be considered an evaluation context in reduction semantics.
We represent frames using the @Frame@ class, which stores the information of a single
call-site, i.e., the information that can be used to resume the interrupted computation.
A @Frame@ constitutes a variable name @x@ to be bound later, a control expression
to which the program may resume, and its environment. \note{JD: someone check this and make sure I got it right}
\todo{mention that frame list vs KArg/KFun}

@@

As previously mentioned, the space of addresses must be finite in AAM.
Binding addresses are parameterized by variable names and the creation time of the binding,
both of which are finite.
Continuation addresses @KAddr@ has two variants:
1) @Halt@ which corresponds to the empty stack, and
2) @ContAddr@ consists of the entry expressions of callee, which are also
finite.
One may ask, why not also keep track of the time in continuation addresses?
\todo{why target expression, allocation polyvariance}

@@

In the previous section, we show that by keeping an unabstracted stack in the state
space, we can recover the precise call/return match.
In this and following sections, we begin describing the transformations step by step.
Our base machine for now is the AAM with unbounded stack, though
as we will later transform the stack to higher-order functions
represent \note{``representing'' or ``represented by''?} continuations, it does not matter what kind of AAM we start from.
This is because transforming to higher-order functions forces us to sequentialize
the order of abstract evaluation, and neither versions of AAM requires construction of a frame of lists \note{should this be frame of lists? sounds weird}, nor to allocate continuations in the store.
Thus, our choice to start from an AAM with unbounded stack is motivated simply because it has an equivalent stack model to abstract definitional interpreters (our final target).

@@

Fusing is a transformation that combines the @step@ and @drive@ functions into a single
function.
The fused function @drive_step@ is essentially formed by merging the functionality of
@step@ into the @drive@ function.
@drive_step@ takes an @NDState@ and a set of explored states as an argument
and returns a set of reachable states once it terminates. \note{Do we really need this section? It seems unlikely that this requires its own section. Could we just append to the last section: For convenience, we combine these functions into one, called \texttt{drivestep}?}

@@

Refunctionalization transforms first-order data types represented as \note{``represented as'' or ``representing?'' It should be representing, right?}
evaluation contexts to higher-order functions.
Besides sequentializing the order of abstract evaluation by functions,
this transformation can be also regarded as allocating continuation functions of
the defined language in the heap of the metalanguage.

@@

Given a set of elements of type @T@, an initial accumulated value of type @S@, a function @f@
that works on element type @T@ and accumulated type @S@, and a function @g@ that
works on and returns a value of type @S@, @nd@ applies @f@ to each element in the
set iteratively, accumulates the value, and finally applies @g@ to it.
Readers familiar with functional programming may think of @fold@ at this point; indeed, it is an implementation of @fold@ in continuation style. But here we have two continuations, one for the next element in the set, and one for the final result.
This is sometimes referred to as an extended continuation-passing style (ECPS) that has a continuation and a meta-continuation \cite{Danvy:1990:AC:91556.91622}.
\todo{correspondence to disentangled version, refunc the NDCont}

@@

This paper is inspired a lot from \citeauthor{darais2017abstracting}'s work.
Instead of just showing the final form of abstract definitional interpreters,
We reveal that refunctionalization plays an important role for inheriting
stack from the meta-language.
The target of the series of transformations in this paper is abstract definitional
interpreters, and we additionally demonstrate abstract interpreters written in both
monadic- and direct-style.
The difference is ours work for A-Normal Form $\lambda$-calculus instead of plain
$\lambda$-calculus, our work also can be easily adopted to plain
$\lambda$-calculus.

>>>>

The abstract syntax in Scala is shown as follows. We assume the source 
program conforms ANF's convention, and not force it in the term structure 
of Scala constructs.

@@

The CESK machine is an
abstract machine for describing semantics of and evaluating $\lambda$-calculus \cite{felleisen1987calculus}.
The CESK machine models program execution as state transitions in a small-step fashion.
As its name suggests, a machine state has four components:
1) \textit{Control} is the expression currently being evaluated.
2) \textit{Environment} is a map that contains the address of a variable in the lexical scope.
3) \textit{Store} models the heap of a program as a map from addresses to values.
The address space consists of numbers (0-indexed).
In our toy language, the only category of value is a closure, i.e., a function paired with an environment.
4) \textit{Continuation} represents the program stack. In this paper, 
we instantiate the stack as a list of frames because the ANF simplifies the evaluation context.
For direct style $\lambda$-calculus, the continuations of CEK/CESK machines can be presented by 
variants of a data type.

The Scala representations for the components of the CESK machine are as follows:

@@

It is worth noting that the continuation class @Kont@ is defined as a list of
frames, where a frame can be considered an evaluation context in reduction semantics.
We represent frames using the @Frame@ class, which stores the information of a single
call-site, i.e., the information that can be used to resume the interrupted computation.
A @Frame@ constitutes a variable name @x@ to be bound later, a control expression
to which the program may resume, and its environment.

@@

As previously mentioned, the space of addresses must be finite in AAM.
Binding addresses are parameterized by variable names and the creation time of the binding,
both of which are finite.
Continuation addresses @KAddr@ has two variants:
1) @Halt@ which corresponds to the empty stack, and
2) @ContAddr@ consists of the target expressions of callee, which are also
finite.
%One may ask, why not also keep track of the time in continuation addresses?
%\todo{why target expression, allocation polyvariance}

@@

In the previous section, we show that by keeping an unabstracted stack in the state
space, we can recover the precise call/return match.
In this and following sections, we begin describing the transformations step by step.
Our base machine for now is the AAM with unbounded stack, though
as we will later transform the stack to higher-order functions
representing continuations, it does not matter what kind of AAM we start from.
This is because transforming to higher-order functions forces us to sequentialize
the order of abstract evaluation, and neither versions of AAM requires construction of a frame of lists \note{should this be frame of lists? sounds weird}, nor to allocate continuations in the store.
Thus, our choice to start from an AAM with unbounded stack is motivated simply because it has an equivalent stack model to abstract definitional interpreters (our final target).

@@

Fusing is a transformation that combines the @step@ and @drive@ functions into a single
function.
The fused function @drive_step@ is essentially formed by merging the functionality of
@step@ into the @drive@ function.
@drive_step@ takes an @NDState@ and a set of explored states as an argument
and returns a set of reachable states once it terminates.

@@

Refunctionalization transforms first-order data types representing
evaluation contexts to higher-order functions.
Besides sequentializing the order of abstract evaluation by functions,
this transformation can be also regarded as allocating continuation functions of
the defined language in the heap of the metalanguage.

@@

Given a set of elements of type @T@, an initial accumulated value of type @S@, a function @f@
that works on element type @T@ and accumulated type @S@, and a function @g@ that
works on and returns a value of type @S@, @nd@ applies @f@ to each element in the
set iteratively, accumulates the value and finally applies @g@ on it.

If we look at @continue@ and @ndcontinue@ in the disentangled AAM, we will find
that @nd@ is an abstraction of those two functions. In function @continue@,
there are two out calls to @ndcontinue@ and @drive_step@, this is the reason why 
we have functions @f@ and @g@ in @nd@.

Readers familiar with functional programming may think of @fold@ at this point; indeed, 
it is an implementation of @fold@ in continuation style. But here we have two 
continuations, one for the next element in the set, and one for the final result.
This is sometimes referred to as an extended continuation-passing style (ECPS) 
that has a continuation and a meta-continuation \cite{Danvy:1990:AC:91556.91622}.

@@

This paper is inspired greatly from \citeauthor{darais2017abstracting}'s work.
Instead of just showing the final form of abstract definitional interpreters,
We reveal that refunctionalization plays an important role for inheriting
stack from the meta-language.
The target of the series of transformations in this paper is abstract definitional
interpreters, and we additionally demonstrate abstract interpreters written in both
monadic- and direct-style.
The difference is ours work for A-Normal Form $\lambda$-calculus instead of plain
$\lambda$-calculus, our work also can be easily adopted to plain
$\lambda$-calculus.
